---
title: "Cohort construction with provenance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort construction with provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cohort` provides the `CohortPipeline` R6 class for building analytic
cohorts with full provenance. Cohort construction is kept strictly upstream
of analysis: the pipeline produces analytic data tables, then hands them
off to whatever consumes them downstream.

## A worked example

We start with a small simulated dataset.

```{r}
library(cohort)
library(data.table)

set.seed(1)
d <- data.table(
  id  = 1:200,
  age = sample(c(NA, 16:80), 200, replace = TRUE),
  sex = sample(c("F", "M", NA), 200, replace = TRUE,
               prob = c(0.48, 0.48, 0.04)),
  grp = sample(c("a", "b"), 200, replace = TRUE)
)
```

### Step 1: install a base table

```{r}
cp <- CohortPipeline$new(
  d,
  cache_file = file.path(tempdir(), "cohort_cache.rds"),
  label      = "Eligible patients"
)
on.exit(cp$save(), add = TRUE)
```

`CohortPipeline$new()` makes a defensive copy of `d` once. The user's
data table is never mutated, no matter what subsequent operations you
run on the pipeline.

The `cache_file` argument enables incremental re-execution: on the
first run a snapshot is written to disk; on subsequent runs the
snapshot is restored and matching operations replay instantly without
recomputation. The `label` argument is the cohort's display label in
CONSORT diagrams (defaults to `"Cohort participants"`); the
identifier `"root"` is universal and always used in code.

### Step 2: apply root-level exclusions

Every exclusion takes a human-readable reason and an R expression as a
string. The string is parsed and evaluated against the rows currently
included on the branch.

```{r}
cp$exclude_and_track("root", "Missing sex", "is.na(sex)")
cp$exclude_and_track("root", "Missing age", "is.na(age)")
cp$exclude_and_track("root", "Under 18",    "age < 18")
```

`NA` predicate results are treated as `FALSE` (rows are kept) so you
can write predicates without defensive `!is.na(...)` clauses for every
column.

### Step 3: branch into sub-cohorts

```{r}
cp$new_cohort("adults_female", from = "root", label = "Adult females")
cp$exclude_and_track("adults_female", "Not female", "sex != 'F'")

cp$new_cohort("adults_grp_a", from = "root",
              label = "Adults, group A")
cp$exclude_and_track("adults_grp_a", "Group not a", "grp != 'a'")
```

Identifiers (`"adults_female"`, `"adults_grp_a"`) are what your code
references; labels are what shows up in figures. Re-issuing
`new_cohort()` with a different label silently updates the label
without invalidating the cache.

A branch starts identical to its parent at the moment of branching.
Sibling branches evolve independently.

### The freeze rule

A cohort becomes **frozen** the first time either:

1. another cohort branches from it, or
2. an artifact is set on it.

After freezing, `$exclude_and_track()` on that cohort errors. The rule
guarantees that a cohort's name maps to one definition forever: once
children depend on it, its exclusion list is fixed. The practical
workflow is "apply all exclusions on a cohort, then branch from it or
attach artifacts." Multi-way forks are unaffected — you can branch a
frozen cohort as many times as you like.

### Step 4: derive cached artifacts

`set_artifact()` is for reusable per-cohort objects (analytic-ready data
tables, summary statistics, baseline tables). Each callback receives a
copy of the included rows and the named list of artifacts already
attached to the cohort.

```{r}
cp$set_artifact("dt_for_analysis",
  from = "adults_female",
  fn = function(dt, sib) {
    dt[, age_group := cut(age,
      breaks = c(18, 30, 50, Inf),
      right = FALSE,
      labels = c("18-29", "30-49", "50+"))]
    dt
  }
)
```

Each `set_artifact()` callback receives a *fresh* copy of the cohort's
included rows — it does not see modifications made by previous
artifacts. To chain derivations, read the previous artifact off the
`sib` argument:

```{r}
cp$set_artifact("baseline_table",
  from = "adults_female",
  fn = function(dt, sib) {
    sib$dt_for_analysis[, .(.N, mean_age = mean(age)),
      by = age_group][order(age_group)]
  }
)
```

### Step 5: inspect the cohort tree

```{r}
print(cp)
cp$list_cohorts()
cp$consort()
```

`$consort()` returns a long-form table — one row per exclusion step,
across every branch. Each branch contributes only its own steps; steps
inherited from a parent at branch time are reported under the parent.

### Step 6: hand off downstream

Cached artifacts are plain R objects — pull them out with
`$get_artifact()` and pass them to whatever consumes the analytic data.

```{r}
analytic_dt <- cp$get_artifact("adults_female", "dt_for_analysis")
baseline    <- cp$get_artifact("adults_female", "baseline_table")

head(analytic_dt)
baseline
```

If you use an analysis-orchestration package, register each artifact as
a named data entry — typically one short loop over `cp$list_artifacts()`.

## Schemas

Use schemas to declare a column-type contract on a branch and verify it
before downstream code consumes the data.

```{r}
cp$declare_schema("adults_female", schema = list(
  age = list(type = "numeric",   na = FALSE),
  sex = list(type = "character", na = FALSE)
))
cp$validate()
```

If a column is missing, has the wrong type, or carries unexpected `NA`s,
`$validate()` throws a single error listing every problem. Pass
`auto_validate = TRUE` to `CohortPipeline$new()` to fail at the failure
site (every `$new_cohort()` and `$set_artifact()` call validates after).

## CONSORT diagrams

The simplest way to draw CONSORT diagrams is `$plot()`. With no
arguments it renders one panel per cohort, walking the root-to-cohort
path; cohort names become box labels.

```{r consort, fig.width = 10, fig.height = 6, fig.align = "center", out.width = "85%"}
cp$plot()
```

You can name specific cohorts:

```{r, fig.width = 7, fig.height = 6, fig.align = "center", out.width = "85%"}
cp$plot(c("adults_female", "adults_grp_a"))
```

Or save to disk:

```r
cp$plot(file = "Figure_1_CONSORT.pdf")
```

For full control over labels, panel titles, side branches and layout,
use `$draw_consort_panels()` (see `?CohortPipeline`).

## How the cache decides what to recompute

The worked example above already uses `cache_file` and `on.exit(cp$save())`.
On the second run of the same script, every operation is checked
against the cached log:

| Method                | Cache key                          |
|-----------------------|------------------------------------|
| `exclude_and_track`   | `(branch, reason, expr_str)`       |
| `new_cohort`          | `(name, from)`                     |
| `declare_schema`      | overwrites; not cached             |
| `set_artifact`        | `(name, from, body(fn), argset)`   |

A match advances the replay cursor with no work done. A mismatch
truncates the log at that point, drops downstream artifacts, cascades
to descendant cohorts whose inherited prefix is now stale, and
recomputes only what changed. Labels are presentation, not part of
the cache key, so re-issuing `new_cohort("adults_female", from =
"root", label = "Adult women")` updates the label without
invalidating anything.

For `set_artifact`, prefer the 3-argument signature
`function(dt, sib, argset)`: data dependencies in `argset` participate
in the cache key, so changing `argset = list(washout = 84L)` to
`washout = 90L` triggers a recompute. The 2-argument form
`function(dt, sib)` still works but won't catch closure-captured
changes.

The cache key uses `body(fn)` literally; if `fn` calls a helper that
you change, the cache cannot detect that. Either include a version tag
in `argset`, or call `cp$invalidate(cohort)` to drop the cohort and
force recompute.

## Mutation contract (summary)

| Method                       | Returns                    | Mutation safe? |
|------------------------------|----------------------------|----------------|
| `$get_included(c)`           | Independent copy           | Yes            |
| Callback in `$set_artifact`  | Independent copy of subset | Yes            |
| `$get_everyone(c)`           | Independent copy + status  | Yes            |

Operations that go through the public API never modify the user's input
data table or another branch.

## Performance notes

- Branching is O(n) in the number of rows of the base table. The data
  values themselves are stored once and shared across the tree.
- `$exclude_and_track()` evaluates the predicate against the included
  subset only; predicates may safely assume earlier exclusions have
  already removed invalid rows.
- The exclusion log is accumulated as a list and materialized as a
  `data.table` only on read (`$consort()`, `$list_cohorts()`), avoiding
  quadratic `rbind` growth.
